---
title: "DIVAS Application Example: GNP Economic Data"
author: "Jiadong Mao, Yinuo Sun"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DIVAS Application Example: GNP Economic Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = FALSE
)
```

# Introduction

In this vignette, we demonstrate how to apply DIVAS to a real-world dataset: the GNP (Gross National Product) economic data. This dataset contains quarterly economic measurements from multiple countries, providing an excellent example for multi-modal data integration.

## Dataset Description

The GNP dataset includes economic measurements from several countries. For our analysis, we'll divide these measurements into three data blocks:

1. **US Economic Indicators**
2. **Japan Economic Indicators**
3. **European Economic Indicators**

This division allows us to explore how economic patterns are shared or unique across different regions.

# Setup

First, let's load the necessary packages:

```{r, eval=FALSE}
# Load DIVAS and other packages
library(DIVAS)
library(qs)       # For loading the .qs data format
library(ggplot2)
library(gridExtra)
library(dplyr)

# Set random seed for reproducibility
set.seed(123)
```

# Data Loading and Preparation

Let's load the GNP dataset and prepare it for DIVAS analysis:

```{r, eval=FALSE}
# Load GNP data
data_path <- system.file("extdata", "gnp_imputed.qs", package = "DIVAS")

# If the package isn't installed with data, use a local path:
# data_path <- "path/to/gnp_imputed.qs"

if(file.exists(data_path)) {
  gnp_data <- qs::qread(data_path)
  
  # Print available variables
  cat("Available variables in the GNP dataset:\n")
  print(colnames(gnp_data))
  
  # Create regional data blocks
  us_vars <- c("US_GNP", "US_CPI", "US_M1", "US_TBILL", "US_UNEMP")
  japan_vars <- c("JP_GNP", "JP_CPI", "JP_M1", "JP_TBILL")
  europe_vars <- c("GER_GNP", "UK_GNP", "FR_GNP", "UK_CPI", "GER_TBILL")
  
  # Create data blocks
  X1 <- as.matrix(gnp_data[, us_vars])
  X2 <- as.matrix(gnp_data[, japan_vars])
  X3 <- as.matrix(gnp_data[, europe_vars])
  
  # Store in list
  datablock <- list(X1 = X1, X2 = X2, X3 = X3)
  
  # Display basic information about each data block
  for (i in 1:length(datablock)) {
    cat(sprintf("Data Block %d (%s): %d x %d\n", 
                i, 
                c("US", "Japan", "Europe")[i],
                nrow(datablock[[i]]), 
                ncol(datablock[[i]])))
  }
} else {
  # Create synthetic economic data if the dataset is not available
  message("GNP dataset not found. Creating synthetic economic data instead...")
  
  # Number of quarters
  n <- 100
  
  # Create timestamps (for plotting)
  quarters <- seq(as.Date("1975-01-01"), by = "quarter", length.out = n)
  
  # Create shared economic cycle
  set.seed(123)
  global_trend <- sin(seq(0, 2*pi*2, length.out = n)) + 
                  0.5*sin(seq(0, 2*pi*8, length.out = n))
  
  # US data (5 variables)
  us_unique <- 0.7*sin(seq(0, 2*pi*1.5, length.out = n))
  X1 <- matrix(0, n, 5)
  for(i in 1:5) {
    X1[,i] <- global_trend + us_unique + 0.3*rnorm(n)
    # Add some variable-specific behavior
    if(i == 2) X1[,i] <- X1[,i] + 0.5*cos(seq(0, 2*pi*3, length.out = n))
    if(i == 4) X1[,i] <- X1[,i] - 0.4*sin(seq(0, 2*pi*5, length.out = n))
  }
  colnames(X1) <- c("US_GNP", "US_CPI", "US_M1", "US_TBILL", "US_UNEMP")
  
  # Japan data (4 variables)
  japan_unique <- 0.6*sin(seq(0, 2*pi*1.2, length.out = n))
  X2 <- matrix(0, n, 4)
  for(i in 1:4) {
    X2[,i] <- global_trend + japan_unique + 0.3*rnorm(n)
    # Add some variable-specific behavior
    if(i == 1) X2[,i] <- X2[,i] + 0.4*cos(seq(0, 2*pi*4, length.out = n))
    if(i == 3) X2[,i] <- X2[,i] - 0.3*sin(seq(0, 2*pi*6, length.out = n))
  }
  colnames(X2) <- c("JP_GNP", "JP_CPI", "JP_M1", "JP_TBILL")
  
  # European data (5 variables)
  europe_unique <- 0.5*sin(seq(0, 2*pi*1.8, length.out = n))
  X3 <- matrix(0, n, 5)
  for(i in 1:5) {
    X3[,i] <- global_trend + europe_unique + 0.3*rnorm(n)
    # Add some variable-specific behavior
    if(i == 2) X3[,i] <- X3[,i] + 0.5*cos(seq(0, 2*pi*2.5, length.out = n))
    if(i == 5) X3[,i] <- X3[,i] - 0.4*sin(seq(0, 2*pi*4.5, length.out = n))
  }
  colnames(X3) <- c("GER_GNP", "UK_GNP", "FR_GNP", "UK_CPI", "GER_TBILL")
  
  # Store in list
  datablock <- list(X1 = X1, X2 = X2, X3 = X3)
  
  # Display basic information about each data block
  for (i in 1:length(datablock)) {
    cat(sprintf("Data Block %d (%s): %d x %d\n", 
                i, 
                c("US", "Japan", "Europe")[i],
                nrow(datablock[[i]]), 
                ncol(datablock[[i]])))
  }
}
```

# Preliminary Exploration

Before running DIVAS, let's briefly explore the data to get a sense of the relationships between variables:

```{r, eval=FALSE}
# Create a function to plot time series from each data block
plot_timeseries <- function(data_block, title, cols = NULL) {
  if(is.null(cols)) cols <- 1:min(3, ncol(data_block))
  
  # Create a data frame for plotting
  df <- data.frame(
    Time = 1:nrow(data_block),
    data_block[, cols, drop = FALSE]
  )
  
  # Reshape for ggplot
  df_long <- tidyr::pivot_longer(df, -Time, 
                                names_to = "Variable", 
                                values_to = "Value")
  
  # Create plot
  p <- ggplot(df_long, aes(x = Time, y = Value, color = Variable)) +
    geom_line() +
    labs(title = title,
         x = "Quarter",
         y = "Value (Standardized)") +
    theme_minimal() +
    theme(legend.position = "bottom")
  
  return(p)
}

# Plot sample time series from each block
p1 <- plot_timeseries(scale(datablock[[1]]), "US Economic Indicators")
p2 <- plot_timeseries(scale(datablock[[2]]), "Japan Economic Indicators")
p3 <- plot_timeseries(scale(datablock[[3]]), "European Economic Indicators")

# Arrange plots
gridExtra::grid.arrange(p1, p2, p3, ncol = 1)
```

# DIVAS Analysis

Now, let's run the DIVAS analysis on our economic data blocks:

```{r, eval=FALSE}
# Run DIVAS analysis with column centering
# Use 500 bootstrap simulations for more stable results
result <- DIVASmain(datablock, nsim = 500, colCent = TRUE)

# Display the estimated rank structure
print(result$estRank)
```

# Results Interpretation

Let's interpret the key findings:

```{r, eval=FALSE}
# Extract information about joint structures
cat("Joint structure summary:\n")
cat("- Structure shared across all three regions:", sum(result$estRank$rank[1,]), "\n")
cat("- Structure shared between US and Japan:", sum(result$estRank$rank[2,]), "\n")
cat("- Structure shared between US and Europe:", sum(result$estRank$rank[3,]), "\n")
cat("- Structure shared between Japan and Europe:", sum(result$estRank$rank[4,]), "\n")
cat("- Structure unique to US:", sum(result$estRank$rank[5,]), "\n")
cat("- Structure unique to Japan:", sum(result$estRank$rank[6,]), "\n")
cat("- Structure unique to Europe:", sum(result$estRank$rank[7,]), "\n")
```

# Visualization

Let's visualize the joint structures identified by DIVAS:

```{r, eval=FALSE}
# Create data block names for plotting
dataname <- c("US", "Japan", "Europe")

# Generate diagnostic plots
plots <- DJIVEAngleDiagnosticJP(datablock, dataname, result, 123, "GNP Economic Data")

# Display the loadings plot
plots$loadings
```

```{r, eval=FALSE}
# Display the scores plot
plots$scores
```

# Analysis of Shared Economic Patterns

The DIVAS analysis has identified shared and unique economic patterns across regions. Let's delve deeper into what these patterns might represent:

## Global Economic Patterns

```{r, eval=FALSE}
# Reconstruct data based on shared structure across all regions
if(exists("result$dimsdir$JCdimdir")) {
  # Extract joint components
  JC_loadings <- result$dimsdir$JCdimdir
  
  # Create a df for the global shared pattern
  if(length(JC_loadings) > 0 && sum(result$estRank$rank[1,]) > 0) {
    # This extracts the shared component across all three blocks
    shared_pattern <- result$dimsdir$JC_score
    
    # Plot the global pattern
    df_global <- data.frame(
      Time = 1:nrow(shared_pattern),
      Pattern = shared_pattern
    )
    
    ggplot(df_global, aes(x = Time, y = Pattern)) +
      geom_line() +
      labs(title = "Global Economic Pattern Shared Across All Regions",
           x = "Quarter",
           y = "Score") +
      theme_minimal()
  } else {
    cat("No significant global pattern found across all three regions.\n")
  }
} else {
  cat("Joint component information not available in the results.\n")
}
```

## Regional Economic Interactions

We can also examine how different pairs of regions share economic patterns:

```{r, eval=FALSE}
# Check if we have pairwise shared patterns
if(exists("result$dimsdir$JPCdimdir")) {
  # Extract pairwise joint components
  JPC_loadings <- result$dimsdir$JPCdimdir
  
  # Check US-Japan shared pattern
  us_japan_idx <- 2  # Index for US-Japan in the structure
  if(sum(result$estRank$rank[us_japan_idx,]) > 0) {
    cat("US and Japan share significant economic patterns.\n")
  } else {
    cat("No significant shared patterns between US and Japan.\n")
  }
  
  # Check US-Europe shared pattern
  us_europe_idx <- 3  # Index for US-Europe in the structure
  if(sum(result$estRank$rank[us_europe_idx,]) > 0) {
    cat("US and Europe share significant economic patterns.\n")
  } else {
    cat("No significant shared patterns between US and Europe.\n")
  }
  
  # Check Japan-Europe shared pattern
  japan_europe_idx <- 4  # Index for Japan-Europe in the structure
  if(sum(result$estRank$rank[japan_europe_idx,]) > 0) {
    cat("Japan and Europe share significant economic patterns.\n")
  } else {
    cat("No significant shared patterns between Japan and Europe.\n")
  }
} else {
  cat("Pairwise joint component information not available in the results.\n")
}
```

# Economic Insights from DIVAS Analysis

The DIVAS analysis of economic data from the US, Japan, and Europe reveals several key insights:

1. **Global Economic Trends**: The shared structure across all three regions suggests the existence of global economic trends that affect all economies, such as global business cycles or international economic shocks.

2. **Regional Dependencies**: The pairwise shared structures indicate stronger economic interdependencies between certain regions, which could be explained by trade relationships, similar monetary policies, or synchronized business cycles.

3. **Regional Uniqueness**: The individual structures unique to each region reflect economic patterns that are driven by region-specific factors, such as domestic policies, regional economic conditions, or structural differences in the economies.

These insights demonstrate the power of DIVAS to uncover meaningful patterns in multi-modal economic data, providing researchers and policymakers with a deeper understanding of how different economies relate to one another.

# Conclusion

In this vignette, we've demonstrated how to apply DIVAS to analyze economic data across different regions. The DIVAS approach enables us to:

1. Identify shared economic patterns across multiple regions
2. Quantify the strength of economic interdependencies between pairs of regions
3. Isolate region-specific economic patterns

This analytical framework is valuable for economic research, policy analysis, and forecasting, as it provides a structured way to decompose complex economic relationships into interpretable components.

The same approach can be extended to other domains, such as multi-omics data in biomedical research, environmental data in climate studies, or social science data in demographic analyses.

# References

1. Prothero, J., et al. (2024). Data integration via analysis of subspaces (DIVAS).
2. Stock, J.H., & Watson, M.W. (2016). Dynamic factor models, factor-augmented vector autoregressions, and structural vector autoregressions in macroeconomics. In Handbook of macroeconomics (Vol. 2, pp. 415-525). Elsevier. 