---
title: "Getting Started with DIVAS"
author: "Yinuo Sun, Jiadong Mao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with DIVAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Highlights

- Multi-modal data integration across different datasets
- Identification of joint structure across data blocks
- Signal extraction and noise filtering
- Visualization of shared patterns across different datasets

# Introduction

DIVAS (Data Integration via Analysis of Subspaces) is a powerful methodology for integrating and analyzing multiple high-dimensional datasets. Based on the approach proposed by Prothero et al. (2024), DIVAS enables researchers to identify shared patterns and structures across diverse data sources.

The method works through three key steps:

1. **Signal Extraction**: Identifying significant signals in each dataset while filtering out noise
2. **Joint Structure Identification**: Finding shared patterns across multiple datasets
3. **Structure Reconstruction**: Reconstructing the original data based on the identified structures

In this vignette, we'll demonstrate how to use DIVAS to analyze a toy dataset with three data blocks.

# Setup

First, let's load the necessary packages:

```{r, eval=FALSE}
# Load DIVAS
library(DIVAS)

# Visualization packages
library(ggplot2)
library(gridExtra)

# Set a random seed for reproducibility
set.seed(566)
```

# Data Preparation

For this example, we'll use a simple toy dataset that demonstrates the DIVAS functionality:

```{r, eval=FALSE}
# Create synthetic data
# Parameters
n <- 100  # Number of samples (rows)
d1 <- 50  # Dimensions for first block (columns)
d2 <- 50  # Dimensions for second block (columns)
d3 <- 50  # Dimensions for third block (columns)

# Create shared structure
shared_score <- matrix(rnorm(n * 2), n, 2)

# Create individual structures
unique_score1 <- matrix(rnorm(n * 1), n, 1)
unique_score2 <- matrix(rnorm(n * 2), n, 2)
unique_score3 <- matrix(rnorm(n * 1), n, 1)

# Create loadings
shared_load1 <- matrix(rnorm(d1 * 2), d1, 2)
shared_load2 <- matrix(rnorm(d2 * 2), d2, 2)
shared_load3 <- matrix(rnorm(d3 * 2), d3, 2)

unique_load1 <- matrix(rnorm(d1 * 1), d1, 1)
unique_load2 <- matrix(rnorm(d2 * 2), d2, 2)
unique_load3 <- matrix(rnorm(d3 * 1), d3, 1)

# Create data blocks with noise
noise_level <- 0.5
X1 <- shared_score %*% t(shared_load1) + unique_score1 %*% t(unique_load1) + 
  noise_level * matrix(rnorm(n * d1), n, d1)
X2 <- shared_score %*% t(shared_load2) + unique_score2 %*% t(unique_load2) + 
  noise_level * matrix(rnorm(n * d2), n, d2)
X3 <- shared_score %*% t(shared_load3) + unique_score3 %*% t(unique_load3) + 
  noise_level * matrix(rnorm(n * d3), n, d3)

# Store in list
datablock <- list(X1 = X1, X2 = X2, X3 = X3)

# Display basic information about each data block
for (i in 1:length(datablock)) {
  cat(sprintf("Data Block %d dimensions: %d x %d\n", 
              i, nrow(datablock[[i]]), ncol(datablock[[i]])))
}
```

# DIVAS Analysis

Now, let's run the DIVAS analysis on our data blocks:

```{r, eval=FALSE}
# Run DIVAS analysis
result <- DIVASmain(datablock, nsim = 300, colCent = TRUE)

# Display the estimated rank structure
print(result$estRank)
```

The output shows the estimated rank for each type of structure:
1. Structure shared across all three blocks
2. Structure shared between blocks 1-2
3. Structure shared between blocks 1-3
4. Structure shared between blocks 2-3
5. Structure unique to block 1
6. Structure unique to block 2
7. Structure unique to block 3

# Visualization

DIVAS provides diagnostic tools to visualize the identified structures:

```{r, eval=FALSE, fig.width=8, fig.height=6}
# Create data block names for plotting
dataname <- paste0("DataBlock_", 1:length(datablock))

# Generate diagnostic plots
plots <- DJIVEAngleDiagnosticJP(datablock, dataname, result, 566, "Toy Data Example")

# Display the loadings plot
plots$loadings
```

```{r, eval=FALSE, fig.width=8, fig.height=6}
# Display the scores plot
plots$scores
```

These diagnostic plots help us understand:

1. **Loadings Plot**: Shows how variables in each data block contribute to the identified structures
2. **Scores Plot**: Shows how samples/observations relate to the identified structures

# Analysis of Results

Let's interpret the key findings:

```{r, eval=FALSE}
# Extract information about joint structures
cat("Joint structure summary:\n")
cat("- Structure shared across all three blocks:", sum(result$estRank$rank[1,]), "\n")
cat("- Structure shared between blocks 1-2:", sum(result$estRank$rank[2,]), "\n")
cat("- Structure shared between blocks 1-3:", sum(result$estRank$rank[3,]), "\n")
cat("- Structure shared between blocks 2-3:", sum(result$estRank$rank[4,]), "\n")
cat("- Structure unique to block 1:", sum(result$estRank$rank[5,]), "\n")
cat("- Structure unique to block 2:", sum(result$estRank$rank[6,]), "\n")
cat("- Structure unique to block 3:", sum(result$estRank$rank[7,]), "\n")
```

# Data Reconstruction

DIVAS can reconstruct the original data blocks based on the identified structures:

```{r, eval=FALSE}
# Reconstruct data blocks
reconstructed <- DJIVEReconstructMJ(datablock, result$dimsdir)

# Calculate reconstruction accuracy
calc_rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

cat("Reconstruction RMSE:\n")
for (i in 1:length(datablock)) {
  rmse <- calc_rmse(datablock[[i]], reconstructed[[i]])
  cat(sprintf("Block %d: %.6f\n", i, rmse))
}
```

The reconstruction error indicates how well the identified structures capture the information in the original data.

# Applications of DIVAS

DIVAS can be applied to many types of multi-modal data:

| Data Type 1 | Data Type 2 | Application |
|-------------|-------------|-------------|
| Genomics    | Transcriptomics | Multi-omics integration |
| Clinical measurements | Molecular data | Biomedical research |
| Environmental variables | Ecological measurements | Environmental science |
| Demographic data | Economic indicators | Social science research |

# Conclusion

In this vignette, we've demonstrated how to use DIVAS to:

1. Load and prepare multi-block data
2. Run the DIVAS analysis
3. Visualize and interpret the results
4. Reconstruct data based on identified structures

DIVAS provides a powerful framework for integrating and analyzing multiple high-dimensional datasets, enabling researchers to discover shared patterns that might not be apparent when analyzing each dataset separately.

# References

Prothero, J., et al. (2024). Data integration via analysis of subspaces (DIVAS). 